\toolnameend: A Tool for Visualizing Patterns of Spreadsheet Function Combinations

### 2016-08-13 doc/paper.tex
Goodbye, middle initial.
Glyph work

	% author names and affiliations
	% use a multiple column layout for up to three different
	% affiliations
	\author{ \IEEEauthorblockN{Justin A. Middleton and Emerson Murphy-Hill}
		\IEEEauthorblockA{North Carolina State University, Raleigh, USA\\ jamiddl2@ncsu.edu, emerson@csc.ncsu.edu }}

@@

	Fortunately, spreadsheet researchers address these dangers by amassing
	spreadsheet collections from diverse sources, industrial to academic, so
	researchers can study how people use spreadsheets. Collections like
	EUSES~\cite{fisher2005euses}, Fuse~\cite{barik2015fuse}, and the Enron
	corpus~\cite{hermans2015enron} have already produced valueable research, such
	as the documentation of code smells in spreadsheet
	environments~\cite{hermans2012detecting}~\cite{jansen2015code} or the creation
	of grammars to describe every possible spreadsheet program in
	Excel~\cite{aivaloglou2015grammar}. Additionally, some research compares these
	collections to discover variation in spreadsheet techniques over time and
	location~\cite{jansen2015enron}.

@@

	Implicit in this study is an exploration of the Excel API and how people use
	it. Previous research, such as the first study on the Enron
	spreadsheets~\cite{hermans2015enron} or in its comparison with the EUSUS
	corpus~\cite{jansen2015enron}, quantifies how often users employ certain
	functions but not how they combine them. Researchers have applied such
	questions to other contexts as well, such as Murphy and colleague's study into
	how Java developers use Eclipse and what commands they use most
	often~\cite{murphy2006java}. Others focus on the documentation of the API
	itself: Robillard and DeLine, for example conducted a series of surveys and
	interviews to diagnose common problems with learning APIs. They found that
	code examples, which is one of the focuses of this tool, is one of the most important
	aids to have in learning a new system~\cite{robillard2011field}. Though this
	paper applies the tool to the Enron corpus specifically, the tool itself is not
	limited to this; it can be readily applied to any collection of Excel
	spreadsheets a user gives it.

@@

	\subsection{Walkthrough of the Tool} \label{sec:walkthrough}Before discussing
	\toolnameposs development, we will give a basic example of how to use
	it. We will work from the visualization built on the Enron
	corpus to answer this question: \textit{ What kinds of functions do people use
		to define the condition in an IF function?}

	In Excel, the IF functions is one of the logical functions: given a condition
	as the first argument that evaluates to either true or false, it takes on the
	value of either the second or third argument respectively. Or, as the
	documentation writes out, ``IF(Something is True, then do something, otherwise
	do something
	else)"\footnote{\url{https://support.office.com/en-us/article/IF-function-69aed7c9-4e8a-4755-a9bc-aa8bbff73be2?ui=en-US\&rs=en-US\&ad=US}}. Furthermore, the third argument is optional; if left out, the function will return a default ``FALSE" value if necessary. Examples of this function in practice can be seen in Figure~\ref{fig:ifexample}.

@@

	The visualization so far comprises two types of nodes. The first are the
	circles \includegraphics{glossary-greenonly}, which represens a discrete
	function in the formula. These are sized according to their frequency relative
	to the number of times the root function appeared. The second kind of node are
	the numbered squares \includegraphics{glossary-blue}, which represent the
	positions of arguments within its parent function. Looking back at the tree, we
	can confirm that IF functions in this dataset had at most three arguments
	passed into them, which corresponds with its specifications in the API.

	Knowing that the first argument of IF contains the conditionals, we click on
	the square labeled ``1" to explore. When the tool finds more than 10 unique
	arguments in a position, it saves space by displaying only the first ten and
	includes an option to display the rest by clicking the arrow
	\includegraphics{glossary-arrow}.

	\begin{figure}[h] \centering \includegraphics[width=0.7\columnwidth]{IFexpand}
		\caption{First argument of the IF tree} \label{fig:expandif} \end{figure}

	In Figure~\ref{fig:expandif}, we see that IF's first arguments are most often comparison operators,
	such as = and $\le$, and boolean-returning functions, like ISNA and AND. Of these,
	simple equality is the most common and ABS is the least. From here, we
	can further explore the common options among these functions. Clicking on the
	``=" node will yield two arguments and expanding each of them will show
	the common values on either side of the equals sign, as in Figure~\ref{fig:fullpic}.

	\begin{figure*}[t] \centering \includegraphics[width=0.8\textwidth]{IFargslabel}
		\caption{A picture of \toolname displaying the common arguments to the IF
			function's first parameter} \centering \label{fig:fullpic} \end{figure*}

@@

	From these images, we can provide a quick glossary of glyphs within the
	visualization to aid interpretation. \begin{itemize} \item
		\includegraphics{glossary-green} - \textbf{Function nodes} are green circles
		labeled with the function name. Their size is determined by their relative
		frequency, scaled logarithmically, to the frequency of the root node in the
		tree. They may have any number of arguments, including zero, as determined by
		the API. Users can select any of these nodes to view the examples from the
		provided dataset that use this function in the node's position in the tree.

		\item \vspace{.25cm}\includegraphics{glossary-leaf} - \textbf{Value nodes} are
		white circles with a generic label. They represent elements of a formula which
		accept no arguments, like references (A1), ranges (A1:A10), numbers, boolean
		values, strings, and errors. However, \toolname treats still functions with
		zero arguments, like NOW, as function nodes. Like function nodes, these also
		contain examples.

		\item  \vspace{.25cm} \includegraphics{glossary-blue} - \textbf{Argument
			nodes} are dark blue squares labeled with a number. The number is the position
		of an argument within the function in the parent node. In Figure
		\ref{fig:expandif} for example, argument node 1 shows every function or value
		that was seen in the first argument.

		\item  \vspace{.25cm} \includegraphics{glossary-solidline} - \textbf{Solid
			lines} represent a function's list of parameters. As such, they connect the
		function nodes to argument nodes.

		\item  \vspace{.25cm} \includegraphics{glossary-fadingline} - \textbf{Fading
			lines} represents how a formula author can use several different possible
		functions or values within a single argument position. As such, they connect
		argument nodes to function or value nodes.

		\item  \vspace{.25cm} \includegraphics{glossary-arrow} - \textbf{Expansion
			arrows} appear within an argument node when there are over 10 possible
		functions or values in an argument position. By default, only the first 10
		will be shown along with this arrow; clicking this will show the rest as well.

		\item  \vspace{.25cm} \includegraphics[scale=.75]{glossary-lightblue} -
		\textbf{Optional argument trees} are variations of trees that show only
		formulae with a specific number of arguments. For example, the function AND
		can take any number of arguments, up to 255. By default, the tree shows data
		from AND functions with any number of arguments, but these trees, which have
		argument nodes of a lighter blue color, can show how AND was used when
		\textit{only} two arguments were passed into it, or \textit{only} four, and so
		on. See Figure~\ref{fig:optional} for how the data between tree types differs.
	\end{itemize}

@@

	\subsection{Implementation Details} The visualization is the product of two
	discrete processes: \begin{itemize} \item \textit{Collection}: Given a set of
		Excel sheets, the tool, written in Java, uses Apache
		POI\footnote{https://poi.apache.org/} to identify and iterate over every cell
		containing a valid formula. Afterward, it calls POI's formula parser to break
		the formula text into an ordered set of individual tokens, which the tool then
		parses into the tree-like form. When all formulae have been analyzed like this,
		it produces JSON files for each top-level function in the set.

>>>>

	% author names and affiliations
	% use a multiple column layout for up to three different
	% affiliations
	\author{ \IEEEauthorblockN{Justin Middleton and Emerson Murphy-Hill}
		\IEEEauthorblockA{North Carolina State University, Raleigh, USA\\ jamiddl2@ncsu.edu, emerson@csc.ncsu.edu }}

@@

	Fortunately, spreadsheet researchers address these dangers by amassing
	spreadsheet collections from diverse sources, industrial to academic, so
	they can study how people use spreadsheets. Collections like
	EUSES~\cite{fisher2005euses}, Fuse~\cite{barik2015fuse}, and the Enron
	corpus~\cite{hermans2015enron} have already produced valueable research, such
	as the documentation of code smells in spreadsheet
	environments~\cite{hermans2012detecting}~\cite{jansen2015code} or the creation
	of grammars to describe every possible spreadsheet program in
	Excel~\cite{aivaloglou2015grammar}. Additionally, some research compares these
	collections to discover variation in spreadsheet techniques over time and
	location~\cite{jansen2015enron}.

@@

	Implicit in this study is an exploration of the Excel API and how people use
	it. Previous research, such as the first study on the Enron
	spreadsheets~\cite{hermans2015enron} or in its comparison with the EUSUS
	corpus~\cite{jansen2015enron}, quantifies how often users employ certain
	functions but not how they combine them. Researchers have applied such
	questions to other contexts as well, such as Murphy and colleague's study into
	how Java developers use Eclipse and what commands they use most
	often~\cite{murphy2006java}. Others focus on the documentation of the API
	itself: Robillard and DeLine, for example, conducted a series of surveys and
	interviews to diagnose common problems with learning APIs. They found that
	code examples, which is one of the focuses of this tool, is one of the most important
	aids to have in learning a new system~\cite{robillard2011field}. Though this
	paper applies the tool to the Enron corpus specifically, the tool itself is not
	limited to this; it can be readily applied to any collection of Excel
	spreadsheets a user gives it.

@@

	\subsection{Walkthrough of the Tool} \label{sec:walkthrough}Before discussing
	\toolnameposs development, we will give a basic example of how to use
	it. We will work from the visualization built on the Enron
	corpus to answer this question: \textit{ What kinds of functions do people use
		to define the condition in an \texttt{IF} function?}

	In Excel, the \texttt{IF} functions is one of the logical functions: given a condition
	as the first argument that evaluates to either true or false, it takes on the
	value of either the second or third argument respectively. Or, as the
	documentation writes out, ``\texttt{IF(Something is True, then do something, otherwise
	do something
	else)}"\footnote{\url{https://support.office.com/en-us/article/IF-function-69aed7c9-4e8a-4755-a9bc-aa8bbff73be2?ui=en-US\&rs=en-US\&ad=US}}. Furthermore, the third argument is optional; if left out, the function will return a default ``FALSE" value if necessary. Examples of this function in practice can be seen in Figure~\ref{fig:ifexample}.

@@

	The visualization so far comprises two types of nodes. The first are the
	circles \includegraphics{glossary-greenonly}, which represents a discrete
	function in the formula. These nodes are sized according to their frequency relative
	to the number of times the root function appeared. The second kind of node are
	the numbered squares \includegraphics{glossary-blue}, which represent the
	positions of arguments within its parent function. Looking back at the tree, we
	can confirm that \texttt{IF} functions in this dataset had at most three arguments
	passed into them, which corresponds with its specifications in the API.

	Knowing that the first argument of \texttt{IF} contains the conditionals, we click on
	the square labeled ``1" to explore. When the tool finds more than 10 unique
	arguments in a position, it saves space by displaying only the first ten and
	includes an option to display the rest by clicking the arrow
	\includegraphics{glossary-arrow}.

	\begin{figure}[h] \centering \includegraphics[width=0.7\columnwidth]{IFexpand}
		\caption{First argument of the \texttt{IF} tree} \label{fig:expandif} \end{figure}

	In Figure~\ref{fig:expandif}, we see that \texttt{IF}'s first arguments are most often comparison operators,
	such as = and $\le$, and boolean-returning functions, like \texttt{ISNA} and \texttt{AND}. Of these,
	simple equality is the most common and ABS is the least. From here, we
	can further explore the common options among these functions. Clicking on the
	``=" node will yield two arguments and expanding each of them will show
	the common values on either side of the equals sign, as in Figure~\ref{fig:fullpic}.

	\begin{figure*}[t] \centering \includegraphics[width=0.8\textwidth]{IFargslabel}
		\caption{A picture of \toolname displaying the common arguments to the \texttt{IF}
			function's first parameter} \centering \label{fig:fullpic} \end{figure*}

@@

	From these images, we provide glossary of glyphs to better interpret the visualizations. \begin{itemize} \item
		\includegraphics{glossary-green} \textbf{Function nodes} are green circles
		labeled with the function name. Their size is determined by their relative
		frequency, scaled logarithmically, to the frequency of the root node in the
		tree. They may have any number of arguments, including zero, as determined by
		the API. Users can select any of these nodes to view the examples from the
		provided dataset that use this function in the node's position in the tree.

		\item \vspace{.25cm}\includegraphics{glossary-leaf} \textbf{Value nodes} are
		white circles with a generic label. They represent elements of a formula that
		accept no arguments, like references (A1), ranges (A1:A10), numbers, boolean
		values, strings, and errors. However, \toolname still treats functions with
		zero arguments, like \texttt{NOW}, as function nodes. Like function nodes,
		these also link to examples.

		\item  \vspace{.25cm} \includegraphics{glossary-blue} \textbf{Argument nodes}
		are dark blue squares labeled with a number. The number is the position of an
		argument within the function of the parent node. In Figure~\ref{fig:expandif}
		for example, argument node 1 shows every function or value that was seen in
		the first argument.

		\item  \vspace{.25cm} \includegraphics{glossary-solidline} \textbf{Solid
			lines} represent a function's list of parameters. As such, they connect the
		function nodes to argument nodes.

		\item  \vspace{.25cm} \includegraphics{glossary-fadingline} \textbf{Fading
			lines} represents how a formula author can choose from several possible
		functions or values within a single argument position. As such, these lines
		connect argument nodes to function or value nodes.

		\item  \vspace{.25cm} \includegraphics{glossary-arrow} \textbf{Expansion
			arrows} appear within an argument node when there are over 10 possible
		functions or values in an argument position. By default, only the first 10
		will be shown along with this arrow; clicking this will show the rest as well.

		\item  \vspace{.25cm}
		$\vcenter{\hbox{\includegraphics[scale=.75]{glossary-lightblue}}}$
		\textbf{Optional argument trees} are variations of trees that show only
		formulae with a specific number of arguments. For example, the function
		\texttt{AND} can take any number of arguments, up to 255. By default, the tree
		shows data from \texttt{AND} functions with any number of arguments, but these
		trees, which have argument nodes of a lighter blue color, can show how
		\texttt{AND} was used when \textit{only} two arguments were passed into it, or
		\textit{only} four, and so on. See Figure~\ref{fig:optional} for how the data
		between tree types differs.
	\end{itemize}

@@

	\subsection{Implementation Details} The visualization is the product of two processes: \begin{itemize} \item \textit{Collection}: Given a set of
		Excel sheets, the tool, written in Java, uses Apache
		POI\footnote{https://poi.apache.org/} to identify and iterate over every cell
		containing a valid formula. Afterward, it calls POI's formula parser to break
		the formula text into an ordered set of individual tokens, which the tool then
		parses into the tree-like form. When all formulae have been analyzed like this,
		it produces JSON files for each top-level function in the set.
